Freezing of a Weeks-Chandler-Anderson Fluid using classical Density Functional Theory

### 2020-09-08 papers/thesis-kirstie/thesis.tex
smoothed some stuff out

Another way to form an expression for the excess Helmholtz free energy in terms of a homogeneous number density is to use Scaled Particle Theory (SPT).
Scaled particle theory seeks a general expression for the excess chemical potential of a homogeneous Hard Sphere fluid which can then be integrated to find the excess Helmholtz free energy.
\begin{equation}F_{ex}=\int{\mu_{ex}dN}\end{equation}
It is an extension of the work done by Widom who showed that the excess chemical potential of a homogenesous fluid corredsponds to the reversible work 
of inserting a sphere of radius R into a fluid consisting of hard spheres of radius R. 
Widom's result was obtained by considering the difference in the excess Helmholtz free energy when one atom is added to an otherwise closed system. 
He showed that the excess chemcial potential is related to the ensemble average of the Boltzman factor for the potential energy difference $V_{diff}$ 
between a system with N atoms and one with N+1 atoms. The resulting equation is known as Widom's Insertion Formula.
%\begin{displaymath}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~~}\end{displaymath}
\begin{equation}\mu_{ex}=\frac{F_{ex}(N+1)-F_{ex}(N)}{\Delta{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\label{widoms-insertion-formula}{~}=-k_BT\ln\left(<e^{-\beta{V_{diff}}}>\right)\end{equation}
\begin{equation}=-k_BT(\ln(1-n\frac{4\pi}{3}(R_o+R)^3))\end{equation}
Forming a more general result for different sizes of cavities in the fluid, Scaled Partical Theory arrives at an expression for the excess Helmholtz free energy of a homogeneous fluid in terms of the packing fraction $\eta$ 
\begin{equation}\label{Fexcess-SPT}{\frac{\beta{F_{excess}}}{N}=-ln(1-\eta)+\frac{3\eta}{1-\eta}+\frac{3{\eta}^2}{2(1-\eta)^2}}\end{equation} 
where
\begin{equation}{\eta = \frac{\mbox{Volume occupied by the spheres}}{\mbox{Total Volume}}}\end{equation}
\begin{equation}{\eta = \frac{N\frac{4}{3}\pi{R^3}}{V}}\end{equation}

@@

\noindent The excess free energy in Eq~\ref{SPT_FE} can be written in terms of the number density $\frac{N}{V}$
\begin{equation}{F_{excess}=k_{B}T\left(-\frac{N}{V}ln(1-\eta)+\frac{N}{V}\frac{3\eta}{1-\eta}+\frac{N}{V}\frac{3{\eta}^2}{2(1-\eta)^2}\right)V}\end{equation}
and using the expressions in Equations~\ref{n3} through~\ref{n0}, this becomes
\begin{equation}\label{FexfromSPT}{F_{excess}=k_{B}T\left(-n_{0}ln(1-n_{3})+\frac{n_{1}n_{2}}{1-n_{3}}+\frac{n_{2}^3}{24\pi(1-n_{3})^2}\right)V}\end{equation}
which has the form 
\begin{equation}{F_{excess} = k_BT(\Phi_1+\Phi_2+\Phi_3)V}\end{equation}
where 
\begin{align}
   \Phi_1 &= -n_{0}ln(1-n_{3}) \\
   \Phi_2 &= \frac{n_{1}n_{2}}{1-n_{3}} \\
   \Phi_3 &= \frac{n_{2}^3}{24\pi(1-n_{3})^2} 
\end{align}

@@

In this paper,  $w_{2}(r)$ will be constructed using the following expresssion for $w_3(r)$ proposed by Schmidt$^{(?)}$ 
\begin{equation}{w_3(r)=\frac{1}{2}}\left[1-\operatorname{erf}\left(\frac{r-\frac{\alpha}{2}}{\frac{\Xi}{\sqrt{2}}}\right)\right]\end{equation} 
 where $w_3(r)$ is subject to the boundary conditions $w_3(0)\approx{1}$ and $w_3(\infty)=0$. 
\color{red}These boundary conditions are satisfied for $\Xi{\sqrt{2}<<\alpha}$, or $k_BT << 2.21$. \color{black} 
Using Equation~\ref{w2_w3_relation}, shown again below, 
\begin{displaymath}{w_{3}=\int_{r}^{\infty}{w_{2}(\vec{r})dr}\mbox{~~~~or equivilantly,~~~~}-\frac{\partial{w_3(r)}}{\partial{r}}=w_2(r)}\end{displaymath}
and solving for $w_2(r)$
\begin{displaymath}{\frac{dw_3(r)}{dr}\bigg|_{\infty}-\frac{dw_3(r)}{dr}\bigg|_{r}=w_2(r)}\end{displaymath} 
we get \begin{equation}{ w_2(r)=\frac{\sqrt{2}}{\Xi\sqrt\pi}\exp^{-\left(\frac{r-\frac{\alpha}{2}}{\Xi/\sqrt{2}}\right)^2} }\end{equation} %From the deconvolution equation \begin{equation}{-\frac{1}{2}f(r)=w_0*w_3 + w_1*w_2 -\vec{w}_{V1}*\vec{w}_{V2}}\end{equation} it follows that $w_2$ must satisfy the relation
Putting $w_2(r)$ into Equation~\ref{mayerder=w2convolution}, shown again below, 
\begin{displaymath}{\frac{df(r)}{dr}=\int_{-\infty}^{\infty}{w_2(r')w_2(r-r')dr'}}\end{displaymath}
gives
\begin{equation}{\frac{df(r)}{dr}=\frac{1}{\Xi\sqrt{\pi}}e^{-\left(\frac{r-\alpha}{\Xi}\right)^2}}\end{equation} 
which can be integrated to obtain the Mayer function
\begin{equation}{f(r)=\int_{\infty}^r{ \frac{1}{\Xi\sqrt{\pi}}e^{-\left(\frac{r'-\alpha}{\Xi}\right)^2}{dr'}}}\end{equation} 
The Mayer function for the error function potential $V_{erf}(r)$ is thus given by
\begin{equation}{f(r)=-\frac{1}{2}\left[1-\operatorname{erf}\left(\frac{r-\alpha}{\Xi}\right)\right]}\end{equation} Combining this result with the definition of the Mayer function \begin{displaymath}f(r)=e^{-\frac{V(r)}{K_BT}}-1\end{displaymath} %This gives an interatomic potential of \begin{equation}{V_{erf}=-k_BT\ln(f(r)-1)}\end{equation} 
gives an expression for the for the error function potential  $V(r)=V_{erf}(r)$ 
\begin{equation}{V_{erf}(r)=-k_BT\ln\left[\frac{1}{2}\left(\operatorname{erf}\left(\frac{r-\alpha}{\Xi}\right)+1\right)\right]}\end{equation} The two parameters $\alpha$ and $\Xi$ are given in equations (3) and (4), and are determined to make $V_{erf}(r)$ match $V_{WCA}(r)$ as closely as possible by compensating for the unwanted temperature dependence that shows up in the error function potential $V_{erf}(r)$.

--------------------------------------------------
%END NEW------------------------------------------

In summary, classical Density Functional Theory (cDFT) used in conjunction with Soft Fundamental Measure Theory (SFMT) provides a powerful tool in modeling the thermodynamic properties of fluids. With SFMT an expression for the excess free energy in terms of the number density can be found simply and directly without taking into account the interaction each sphere, or atom, has with every other sphere. For a spatially varying number density, as would be expected for a crystal, the free energy can then be written as a functional of the number density profile $n(\vec{r})$ 
\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation} 
To implement cDFT, the number density profile $n(\vec{r})$ is varied until the free energy is minimized. The minimum free energy and its corresponding number density profile are then the equilibrium values for that temperature. \color{red} $\leftarrow$CAREFUL! \color{black} Thus to use cDFT all that is needed are number density profiles, such as Gaussian profiles that depend on position and which can be varied, for example, by changing the width of the Gaussian. 
\begin{equation}{n(r)=\frac{1}{\left(\sqrt{2\pi}\sigma\right)^3}\int\exp^{-\frac{|\vec{r}-\vec{R}|^2}{2\sigma^2}}d\vec{r}}\end{equation}

For a number density n that is constant everywhere, as would be expected for a homogeneous fluid in equilbrium, the free energy can be written as a function of n where the ideal Helmholtz free energy can be determined analytically, and SFMT is used once again to form an expresssion for the excess Helmholtz free energy
\begin{equation}{F(n)=F_{ideal}(n) + F_{excess}(n)}\end{equation}

>>>>

Another way to form an expression for the excess Helmholtz free energy in terms of a homogeneous number density is to use Scaled Particle Theory (SPT).
Scaled particle theory seeks a general expression for the excess chemical potential of a homogeneous Hard Sphere fluid which can then be integrated to find the excess Helmholtz free energy.
\begin{equation}F_{ex}=\int{\mu_{ex}dN}\end{equation}
It is an extension of the work done by Widom who showed that the excess chemical potential of a homogenesous fluid corredsponds to the reversible work 
of inserting a sphere of radius R into a fluid consisting of hard spheres of radius R. 
Widom's result was obtained by considering the difference in the excess Helmholtz free energy when one atom is added to an otherwise closed system. 
He showed that the excess chemcial potential is related to the ensemble average of the Boltzman factor for the potential energy difference $V_{diff}$ 
between a system with N atoms and one with N+1 atoms. The resulting equation is known as Widom's Insertion Formula.
%\begin{displaymath}\mu_{ex}=\frac{\partial{F_{ex}}}{\partial{N}}\bigg|_{T,V}{~~~~~~}\end{displaymath}
\begin{equation}\mu_{ex}=\frac{F_{ex}(N+1)-F_{ex}(N)}{\Delta{N}}\bigg|_{T,V}\end{equation}
\begin{equation}\label{widoms-insertion-formula}{~}=-k_BT\ln\left(<e^{-\beta{V_{diff}}}>\right)\end{equation}
\begin{equation}=-k_BT(\ln(1-n\frac{4\pi}{3}(R_{cavity}+R)^3))\end{equation}
where the radius of the cavity $R_{cavity}$ is equal to the radius of a hard sphere, R. Forming a more general result for different sizes of cavities in the fluid, Scaled Partical Theory arrives at an expression for the excess Helmholtz free energy of a homogeneous fluid in terms of the packing fraction $\eta$ 
\begin{equation}\label{Fexcess-SPT}{\frac{\beta{F_{excess}}}{N}=-ln(1-\eta)+\frac{3\eta}{1-\eta}+\frac{3{\eta}^2}{2(1-\eta)^2}}\end{equation} 
where
\begin{equation}{\eta = \frac{\mbox{Volume occupied by the spheres}}{\mbox{Total Volume}}}\end{equation}
\begin{equation}{\eta = \frac{N\frac{4}{3}\pi{R^3}}{V}}\end{equation}

@@

\noindent The excess free energy in Eq~\ref{Fexcess-SPT} can be written in terms of the number density $\frac{N}{V}$
\begin{equation}{F_{excess}=k_{B}T\left(-\frac{N}{V}ln(1-\eta)+\frac{N}{V}\frac{3\eta}{1-\eta}+\frac{N}{V}\frac{3{\eta}^2}{2(1-\eta)^2}\right)V}\end{equation}
and using the expressions in Equations~\ref{n3} through~\ref{n0}, this becomes
\begin{equation}\label{FexfromSPT}{F_{excess}=k_{B}T\left(-n_{0}ln(1-n_{3})+\frac{n_{1}n_{2}}{1-n_{3}}+\frac{n_{2}^3}{24\pi(1-n_{3})^2}\right)V}\end{equation}
which has the form 
\begin{equation}{F_{excess} = k_BT(\Phi_1+\Phi_2+\Phi_3)V}\end{equation}
where 
\begin{align}
   \Phi_1 &= -n_{0}ln(1-n_{3}) \\
   \Phi_2 &= \frac{n_{1}n_{2}}{1-n_{3}} \\
   \Phi_3 &= \frac{n_{2}^3}{24\pi(1-n_{3})^2} 
\end{align}

@@

In this paper,  $w_{2}(r)$ will be constructed using the following expresssion for $w_3(r)$ proposed by Schmidt$^{(?)}$ 
\begin{equation}{w_3(r)=\frac{1}{2}}\left[1-\operatorname{erf}\left(\frac{r-\frac{\alpha}{2}}{\frac{\Xi}{\sqrt{2}}}\right)\right]\end{equation} 
 where $w_3(r)$ is subject to the boundary conditions $w_3(0)\approx{1}$ and $w_3(\infty)=0$. 
%\color{red}These boundary conditions are satisfied for $\Xi{\sqrt{2}<<\alpha}$, or $k_BT << 2.21$. \color{black} 
Using Equation~\ref{w2_w3_relation}, shown again below, 
\begin{displaymath}{w_{3}=\int_{r}^{\infty}{w_{2}(\vec{r})dr}\mbox{~~~~or equivilantly,~~~~}-\frac{\partial{w_3(r)}}{\partial{r}}=w_2(r)}\end{displaymath}
and solving for $w_2(r)$
\begin{displaymath}{\frac{dw_3(r)}{dr}\bigg|_{\infty}-\frac{dw_3(r)}{dr}\bigg|_{r}=w_2(r)}\end{displaymath} 
we get \begin{equation}{ w_2(r)=\frac{\sqrt{2}}{\Xi\sqrt\pi}\exp^{-\left(\frac{r-\frac{\alpha}{2}}{\Xi/\sqrt{2}}\right)^2} }\end{equation} %From the deconvolution equation \begin{equation}{-\frac{1}{2}f(r)=w_0*w_3 + w_1*w_2 -\vec{w}_{V1}*\vec{w}_{V2}}\end{equation} it follows that $w_2$ must satisfy the relation
Putting $w_2(r)$ into Equation~\ref{mayerder=w2convolution}, shown again below, 
\begin{displaymath}{\frac{df(r)}{dr}=\int_{-\infty}^{\infty}{w_2(r')w_2(r-r')dr'}}\end{displaymath}
gives
\begin{equation}{\frac{df(r)}{dr}=\frac{1}{\Xi\sqrt{\pi}}e^{-\left(\frac{r-\alpha}{\Xi}\right)^2}}\end{equation} 
which can be integrated to obtain the Mayer function
\begin{equation}{f(r)=\int_{\infty}^r{ \frac{1}{\Xi\sqrt{\pi}}e^{-\left(\frac{r'-\alpha}{\Xi}\right)^2}{dr'}}}\end{equation} 
The Mayer function for the error function potential $V_{erf}(r)$ is thus given by
\begin{equation}{f(r)=-\frac{1}{2}\left[1-\operatorname{erf}\left(\frac{r-\alpha}{\Xi}\right)\right]}\end{equation} Combining this result with the definition of the Mayer function \begin{displaymath}f(r)=e^{-\frac{V(r)}{K_BT}}-1\end{displaymath} %This gives an interatomic potential of \begin{equation}{V_{erf}=-k_BT\ln(f(r)-1)}\end{equation} 
gives an expression for the for the error function potential  $V(r)=V_{erf}(r)$ 
\begin{equation}{V_{erf}(r)=-k_BT\ln\left[\frac{1}{2}\left(\operatorname{erf}\left(\frac{r-\alpha}{\Xi}\right)+1\right)\right]}\end{equation} The two parameters $\alpha$ and $\Xi$ are given in equations (3) and (4), and are determined to make $V_{erf}(r)$ match $V_{WCA}(r)$ as closely as possible by compensating for the unwanted temperature dependence that shows up in the error function potential $V_{erf}(r)$.

In summary, classical Density Functional Theory (cDFT) is used in conjunction with Soft Fundamental Measure Theory (SFMT) to find the equilibrium free energy of a system, and the corresponding equilibrium number density profile. In particular, SFMT is used to form the expression for the excess free energy without having to take into account the interaction each sphere, or atom, has with every other sphere. The excess free energy is then added to the ideal free energy calculated analytically to find the resulting Free energy for the system. 
\begin{equation}{F=F_{ideal} + F_{excess}}\end{equation} 
For a spatially varying number density, as would be expected for a crystal, the free energy is a functional of the number density profile $n(\vec{r})$  
\begin{equation}{F[n(\vec{r})]=F_{ideal}[n(\vec{r})] + F_{excess}[n(\vec{r})]}\end{equation} To implement cDFT, the number density profile $n(\vec{r})$ is varied with the temperature fixed until the free energy is minimized. The minimum free energy and its corresponding number density profile are then the equilibrium values for that temperature.
